---
title: "Heart Failure Clinical Records - MLP Analysis"
format:
  pdf:
    embed-resources: true
    code-fold: false
    number-sections: true
toc: true
toc-title: Table of contents
toc-depth: 4
toc-location: left
execute: 
  eval: false
  echo: true
output:
  pdf_document:
  output-file: output.pdf
---

# Introduction
Heart failure is among the causes of mortality in the world and it gives significant obstacles
to clinical prognosis and management of the patients. The ability to predict patient survival
through clinical pointers would enable medical practitioners to make optimal treatment
decisions and calculate high-risk patients earlier.
The Heart Failure Clinical Records data have 299 patients and 12 clinical features such as
age, ejection fraction, serum creatinine and various binary variables like diabetes, anemia and
high blood pressure. The variable of interest, or target variable, is DEATH_EVENT, which is
whether a patient has survived or died by the period of follow-up. Given that the dataset
contains no missing data and the outcome variable is binary, it is highly applicable to binary
classification in a clinical prediction context.
This paper examines the performance of various machine learning models on this dataset,
specifically neural networks. In particular, the paper canl analyze how model architecture,
activation functions and network capacity affect learning dynamics and predictive
performance. A linear logistic regression model is used as a baseline, and multilayer
perceptrons (MLPs) are evaluated to assess the impact of nonlinear representations.

# Heart Failure Clinical Records – MLP Analysis
# Tasks 1–4: Logistic Regression vs MLP, Activation Comparison,
#             Capacity & Regularisation, Responsible Evaluation

# ── 0. SETUP ─────────────────────────────────────────────────────────────────

```{r}
# Lines 9-18
install.packages("caret", dependencies = TRUE)
install.packages("lava")
suppressPackageStartupMessages({
  library(nnet)       # nnet() for MLP, multinom(); single-hidden-layer MLPs
  library(ggplot2)
  library(dplyr)
  library(tidyr)
  library(caret)
  library(pROC)
})
set.seed(42)
```

# Output directory for plots

```{r}
# Lines 21
dir.create("outputs", showWarnings = FALSE)
```

# ── 1. DATA LOADING & PRE-PROCESSING ─────────────────────────────────────────

```{r}
# Lines 25-31
raw <- read.csv("heart_failure_clinical_records_dataset.csv",
                stringsAsFactors = FALSE)
cat("=== Dataset dimensions:", nrow(raw), "rows ", ncol(raw), "cols\n")
cat("=== Target distribution:\n")
print(table(raw$DEATH_EVENT))
cat("=== Missing values:", sum(is.na(raw)), "\n\n")
```
# Task 1: Logistic Regression vs Multilayer Perceptron
In this task, two models, a logistic regression model and a multilayer perceptron (MLP) with
a single hidden layer, were trained to indicate patient survival. Stratified sampling was used
to split the dataset into training, validation, and test sets, thereby maintaining the class
proportions of the survival outcome. The data division was 60% training, 20% validation, and
20% test. All numerical characteristics were standardised to a mean of zero and unit variance
to enhance the optimisation stability of gradient-based models.
Accuracy and ROC-AUC were used to test the performance of both models. Accuracy
quantifies the proportion of correct predictions, and ROC-AUC quantifies how well the
model distinguishes between the two classes across different classification thresholds. A
validation set was used to control the model&#39;s behaviour during training, and the test set was
used only to evaluate the model at the end.

# Separate features and target

```{r}
# Lines 34-36
feature_cols <- setdiff(names(raw), "DEATH_EVENT")
X_raw <- raw[, feature_cols]
y     <- factor(raw$DEATH_EVENT, levels = c(0, 1), labels = c("Survived", "Died"))
```

# Standardise all numeric features (mean=0, sd=1) for stable gradient descent

```{r}
# Lines 39
X_scaled <- as.data.frame(scale(X_raw))
```

# Stratified train / validation / test split (60 / 20 / 20)

```{r}
# Lines 42-56
set.seed(123)

# indices for each class
idx0 <- which(y == 0)
idx1 <- which(y == 1)

# shuffle
idx0 <- sample(idx0)
idx1 <- sample(idx1)

# split function
split_class <- function(idx){
  n <- length(idx)
  train <- idx[1:round(0.6*n)]
  val   <- idx[(round(0.6*n)+1):round(0.8*n)]
  test  <- idx[(round(0.8*n)+1):n]
  list(train=train, val=val, test=test)
}

s0 <- split_class(idx0)
s1 <- split_class(idx1)

train_idx <- c(s0$train, s1$train)
val_idx   <- c(s0$val, s1$val)
test_idx  <- c(s0$test, s1$test)

X_train <- X_scaled[train_idx, ]; y_train <- y[train_idx]
X_val   <- X_scaled[val_idx, ];   y_val   <- y[val_idx]
X_test  <- X_scaled[test_idx, ];  y_test  <- y[test_idx]
```

# Helper: combine features + target into a data frame for nnet

```{r}
# Lines 59-62
make_df <- function(X, y) { df <- X; df$DEATH_EVENT <- y; df }
df_train <- make_df(X_train, y_train)
df_val   <- make_df(X_val,   y_val)
df_test  <- make_df(X_test,  y_test)
```

# Helper: evaluate binary classifier (handles glm and nnet)

```{r}
# Lines 65-83
evaluate_model <- function(model, X, y_true, threshold = 0.5) {
  if (inherits(model, "glm")) {
    prob <- predict(model, newdata = X, type = "response")
  } else {
    raw <- predict(model, newdata = X, type = "raw")
    if (is.matrix(raw) && ncol(raw) >= 2) {
      prob <- raw[, "Died"]
    } else {
      prob <- as.vector(raw)   # nnet binary: single column = P(class2 = Died)
    }
  }
  pred  <- factor(ifelse(prob >= threshold, "Died", "Survived"),
                  levels = c("Survived", "Died"))
  acc   <- mean(pred == y_true)
  cm    <- confusionMatrix(pred, y_true, positive = "Died")
  roc_obj <- roc(as.numeric(y_true == "Died"), prob, quiet = TRUE)
  auc_val <- as.numeric(auc(roc_obj))
  list(accuracy = acc, auc = auc_val, cm = cm, prob = prob, roc = roc_obj)
}
```

# ── 2. TASK 1 – LOGISTIC REGRESSION vs MLP (1 hidden layer, ReLU approx) ─────

```{r}
# Lines 87-89
cat("══════════════════════════════════════════════════════════════\n")
cat("  TASK 1: Logistic Regression  vs  MLP (1 hidden layer)\n")
cat("══════════════════════════════════════════════════════════════\n\n")
```

# ── 2a. Logistic Regression (no hidden layer) ────────────────────────────────
# nnet with size=0 is equivalent to multinomial logistic regression.
# The decision boundary is: P(y=1) = σ(w^T x + b) — a purely linear mapping
# through the sigmoid, so the model cannot represent any curved decision surface.

# Logistic regression via glm — the canonical linear model
# Decision rule: P(y=1|x) = σ(w^T x + b),  boundary = hyperplane

```{r}

set.seed(123)

n <- nrow(raw)

# random shuffle
indices <- sample(1:n)

train_size <- round(0.6 * n)
val_size   <- round(0.2 * n)

train_idx <- indices[1:train_size]
val_idx   <- indices[(train_size + 1):(train_size + val_size)]
test_idx  <- indices[(train_size + val_size + 1):n]

# create splits
X_train <- X_scaled[train_idx, ]
y_train <- y[train_idx]

X_val <- X_scaled[val_idx, ]
y_val <- y[val_idx]

X_test <- X_scaled[test_idx, ]
y_test <- y[test_idx]
```

# Wrap into an evaluate-compatible object using a predict helper

```{r}
# Logistic Regression Model
lr_glm <- glm(
  y_train ~ .,
  data = data.frame(X_train, y_train),
  family = binomial(link = "logit")
)
lr_model <- lr_glm   # alias kept for downstream code
lr_train <- evaluate_model(lr_model, X_train, y_train)
lr_val   <- evaluate_model(lr_model, X_val,   y_val)
cat("── Logistic Regression ──\n")
cat(sprintf("  Train  Acc: %.4f  AUC: %.4f\n", lr_train$accuracy, lr_train$auc))
cat(sprintf("  Val    Acc: %.4f  AUC: %.4f\n\n", lr_val$accuracy, lr_val$auc))
```

# ── 2b. MLP with 1 hidden layer (sigmoid activations) ─────────────────────
# nnet uses logistic (sigmoid) hidden units. We track training dynamics
# epoch-by-epoch using warm-start (Wts carries weights forward each step).

```{r}
# Lines 116-124
n_hidden_task1 <- 16
n_epochs       <- 200
train_acc_mlp  <- numeric(n_epochs)
val_acc_mlp    <- numeric(n_epochs)
train_auc_mlp  <- numeric(n_epochs)
val_auc_mlp    <- numeric(n_epochs)
train_loss_mlp <- numeric(n_epochs)
val_loss_mlp   <- numeric(n_epochs)
```

# Cross-entropy loss for binary nnet (single P(Died) output)

```{r}
# Lines 127-134
xent_loss_bin <- function(prob_died, y_true) {
  eps    <- 1e-9
  labels <- as.character(y_true)
  ll     <- ifelse(labels == "Died",
                   log(prob_died + eps),
                   log(1 - prob_died + eps))
  -mean(ll)
}
```

# Helper: get P(Died) from nnet predict (1-col binary output)

```{r}
# Lines 137
get_died_prob <- function(p) as.vector(p)
```

# First fit to obtain initial weight vector

```{r}
# Lines 140-164
mlp_fit0 <- nnet(DEATH_EVENT ~ ., data = df_train,
                 size=n_hidden_task1, decay=1e-4, maxit=1,
                 trace=FALSE, MaxNWts=5000)
mlp_wts <- mlp_fit0$wts
for (ep in seq_len(n_epochs)) {
  mlp_fit <- nnet(DEATH_EVENT ~ ., data = df_train,
                  size=n_hidden_task1, decay=1e-4, maxit=1,
                  Wts=mlp_wts, trace=FALSE, MaxNWts=5000)
  mlp_wts <- mlp_fit$wts
  p_train_died <- get_died_prob(predict(mlp_fit, newdata=X_train, type="raw"))
  p_val_died   <- get_died_prob(predict(mlp_fit, newdata=X_val,   type="raw"))
  train_loss_mlp[ep] <- xent_loss_bin(p_train_died, y_train)
  val_loss_mlp[ep]   <- xent_loss_bin(p_val_died,   y_val)
  train_acc_mlp[ep]  <- mean((p_train_died >= 0.5) == (y_train == "Died"))
  val_acc_mlp[ep]    <- mean((p_val_died   >= 0.5) == (y_val   == "Died"))
  roc_tr <- roc(as.numeric(y_train=="Died"), p_train_died, quiet=TRUE)
  roc_vl <- roc(as.numeric(y_val  =="Died"), p_val_died,   quiet=TRUE)
  train_auc_mlp[ep]  <- as.numeric(auc(roc_tr))
  val_auc_mlp[ep]    <- as.numeric(auc(roc_vl))
}
```

# Store final weights

```{r}
# Lines 167-174
mlp_model_task1 <- mlp_fit
mlp_train <- evaluate_model(mlp_model_task1, X_train, y_train)
mlp_val   <- evaluate_model(mlp_model_task1, X_val,   y_val)
cat("── MLP (1 hidden layer, 16 units, sigmoid) ──\n")
cat(sprintf("  Train  Acc: %.4f  AUC: %.4f\n", mlp_train$accuracy, mlp_train$auc))
cat(sprintf("  Val    Acc: %.4f  AUC: %.4f\n\n", mlp_val$accuracy, mlp_val$auc))
```

# ── Plot learning curves (Task 1) ────────────────────────────────────────────

```{r}
# Lines 178-201
task1_df <- data.frame(
  epoch     = rep(1:n_epochs, 4),
  value     = c(train_acc_mlp, val_acc_mlp, train_loss_mlp, val_loss_mlp),
  split     = rep(c("Train","Validation","Train","Validation"), each = n_epochs),
  metric    = rep(c("Accuracy","Accuracy","Loss (Cross-Entropy)","Loss (Cross-Entropy)"),
                  each = n_epochs)
)
p_task1 <- ggplot(task1_df, aes(x = epoch, y = value, colour = split, linetype = split)) +
  geom_line(size = 0.8) +
  facet_wrap(~metric, scales = "free_y") +
  scale_colour_manual(values = c("Train" = "#2166ac", "Validation" = "#d6604d")) +
  labs(title    = "Task 1 – MLP Learning Curves (1 Hidden Layer, 16 Units)",
       subtitle = "Sigmoid activations | decay = 1e-4 | 200 epochs",
       x = "Epoch", y = "Metric Value",
       colour = "Split", linetype = "Split") +
  theme_bw(base_size = 13) +
  theme(legend.position = "bottom",
        plot.title    = element_text(face = "bold"),
        strip.background = element_rect(fill = "#f0f0f0"))
ggsave("outputs/task1_learning_curves.png", p_task1,
       width = 10, height = 5, dpi = 150)
cat("  [Saved] outputs/task1_learning_curves.png\n\n")
```
# Task 2: Activation Function Comparison
The influence of various activation functions on the training behaviour of the neural networks
was examined in this task. There were two functions studied, specifically, Rectified Linear
Unit (ReLU) and hyperbolic tangent (tanh). The two models shared the same network
architecture and hyperparameters, including one hidden layer with 16 units, the same learning
rate and regularisation strength, and the same number of training epochs. This ensured that
any performance variations were explained by the selected activation function.

# ── 3. TASK 2 – ACTIVATION FUNCTION COMPARISON (ReLU-like vs tanh) ───────────

```{r}
# Lines 205-207
cat("══════════════════════════════════════════════════════════════\n")
cat("  TASK 2: Activation Function Comparison\n")
cat("══════════════════════════════════════════════════════════════\n\n")
```

# NOTE: nnet's hidden units use the *logistic* (sigmoid) function internally,
# not a user-selectable activation.  To faithfully compare activation functions
# we implement a lightweight, manual single-hidden-layer MLP in pure R so that
# we can swap activations at will.

# ── Manual MLP helpers ───────────────────────────────────────────────────────

```{r}
# Lines 216-229
sigmoid   <- function(x) 1 / (1 + exp(-x))
d_sigmoid <- function(x) { s <- sigmoid(x); s * (1 - s) }
relu      <- function(x) pmax(0, x)
d_relu    <- function(x) as.numeric(x > 0)
tanh_fn   <- function(x) tanh(x)
d_tanh    <- function(x) 1 - tanh(x)^2
softmax   <- function(z) {
  z  <- z - max(z)
  ez <- exp(z)
  ez / sum(ez)
}
```

# Xavier initialisation

```{r}
# Lines 232-235
xavier_init <- function(fan_in, fan_out) {
  limit <- sqrt(6 / (fan_in + fan_out))
  matrix(runif(fan_in * fan_out, -limit, limit), nrow = fan_in)
}
```

# Forward pass: returns list(h_pre, h, o_pre, probs)

```{r}
# Lines 238-244
forward <- function(x, W1, b1, W2, b2, act_fn) {
  h_pre <- W1 %*% x + b1
  h     <- act_fn(h_pre)
  o_pre <- W2 %*% h + b2
  probs <- softmax(as.vector(o_pre))
  list(h_pre = h_pre, h = h, o_pre = o_pre, probs = probs)
}
```

# Cross-entropy loss over mini-batch (matrix X: features x samples)

```{r}
# Lines 247-256
batch_loss <- function(X_mat, y_int, W1, b1, W2, b2, act_fn, lambda = 1e-4) {
  n   <- ncol(X_mat)
  loss <- 0
  for (i in seq_len(n)) {
    fp    <- forward(X_mat[, i], W1, b1, W2, b2, act_fn)
    loss  <- loss - log(fp$probs[y_int[i]] + 1e-9)
  }
  loss / n +
    lambda / 2 * (sum(W1^2) + sum(W2^2))   # L2 regularisation term
}
```

# Batch accuracy

```{r}
# Lines 259-265
batch_acc <- function(X_mat, y_int, W1, b1, W2, b2, act_fn) {
  preds <- sapply(seq_len(ncol(X_mat)), function(i) {
    fp <- forward(X_mat[, i], W1, b1, W2, b2, act_fn)
    which.max(fp$probs)
  })
  mean(preds == y_int)
}
```

# Mini-batch SGD training; returns list of per-epoch metrics

```{r}
# Lines 268-351
train_manual_mlp <- function(X_tr, y_tr, X_vl, y_vl,
                             n_hidden  = 16,
                             act_fn    = relu,
                             d_act_fn  = d_relu,
                             lr        = 0.01,
                             lambda    = 1e-4,
                             n_epochs  = 200,
                             batch_size = 32,
                             seed      = 42) {
  set.seed(seed)
  n_features <- ncol(X_tr)
  n_classes  <- 2L
  ## Transpose to features × samples
  Xtr <- t(as.matrix(X_tr))
  Xvl <- t(as.matrix(X_vl))
  ytr <- as.integer(y_tr)   # 1 = Survived, 2 = Died
  yvl <- as.integer(y_vl)
  ## Initialise weights
  W1 <- xavier_init(n_hidden, n_features)
  b1 <- matrix(0, n_hidden, 1)
  W2 <- xavier_init(n_classes, n_hidden)
  b2 <- matrix(0, n_classes, 1)
  n   <- ncol(Xtr)
  metrics <- data.frame(
    epoch    = integer(n_epochs),
    tr_loss  = numeric(n_epochs),
    vl_loss  = numeric(n_epochs),
    tr_acc   = numeric(n_epochs),
    vl_acc   = numeric(n_epochs)
  )
  for (ep in seq_len(n_epochs)) {
    ## Shuffle
    perm <- sample(n)
    Xs   <- Xtr[, perm, drop = FALSE]
    ys   <- ytr[perm]
    ## Mini-batch SGD
    starts <- seq(1, n, by = batch_size)
    for (s in starts) {
      ends   <- min(s + batch_size - 1, n)
      Xb     <- Xs[, s:ends, drop = FALSE]
      yb     <- ys[s:ends]
      nb     <- ncol(Xb)
      dW1 <- matrix(0, nrow(W1), ncol(W1))
      db1 <- matrix(0, nrow(b1), 1)
      dW2 <- matrix(0, nrow(W2), ncol(W2))
      db2 <- matrix(0, nrow(b2), 1)
      for (i in seq_len(nb)) {
        fp <- forward(Xb[, i], W1, b1, W2, b2, act_fn)
        ## Output layer delta
        delta2              <- fp$probs
        delta2[yb[i]]       <- delta2[yb[i]] - 1   # cross-entropy + softmax gradient
        ## Hidden layer delta
        dh     <- t(W2) %*% delta2
        delta1 <- dh * d_act_fn(fp$h_pre)
        dW2 <- dW2 + delta2 %*% t(fp$h)
        db2 <- db2 + delta2
        dW1 <- dW1 + delta1 %*% t(Xb[, i, drop = FALSE])
        db1 <- db1 + delta1
      }
      ## Average gradients + L2
      W1 <- W1 - lr * (dW1 / nb + lambda * W1)
      b1 <- b1 - lr *  db1 / nb
      W2 <- W2 - lr * (dW2 / nb + lambda * W2)
      b2 <- b2 - lr *  db2 / nb
    }
    metrics$epoch[ep]   <- ep
    metrics$tr_loss[ep] <- batch_loss(Xtr, ytr, W1, b1, W2, b2, act_fn, lambda)
    metrics$vl_loss[ep] <- batch_loss(Xvl, yvl, W1, b1, W2, b2, act_fn, lambda)
    metrics$tr_acc[ep]  <- batch_acc(Xtr, ytr, W1, b1, W2, b2, act_fn)
    metrics$vl_acc[ep]  <- batch_acc(Xvl, yvl, W1, b1, W2, b2, act_fn)
  }
  list(W1=W1, b1=b1, W2=W2, b2=b2, metrics=metrics)
}
```

# ── Train with ReLU ──────────────────────────────────────────────────────────

```{r}
# Lines 354-357
cat("  Training ReLU MLP...\n")
relu_result <- train_manual_mlp(X_train, y_train, X_val, y_val,
                                n_hidden = 16, act_fn = relu, d_act_fn = d_relu,
                                lr = 0.01, n_epochs = 200)
```

# ── Train with tanh ──────────────────────────────────────────────────────────

```{r}
# Lines 360-363
cat("  Training tanh MLP...\n")
tanh_result <- train_manual_mlp(X_train, y_train, X_val, y_val,
                                n_hidden = 16, act_fn = tanh_fn, d_act_fn = d_tanh,
                                lr = 0.01, n_epochs = 200)
```

# ── Plot Task 2 ──────────────────────────────────────────────────────────────

```{r}
# Lines 366-417
build_act_df <- function(metrics, label) {
  data.frame(
    epoch      = rep(metrics$epoch, 4),
    value      = c(metrics$tr_loss, metrics$vl_loss, metrics$tr_acc, metrics$vl_acc),
    split      = rep(c("Train","Validation","Train","Validation"), each = nrow(metrics)),
    metric     = rep(c("Loss","Loss","Accuracy","Accuracy"), each = nrow(metrics)),
    activation = label
  )
}
task2_df <- rbind(
  build_act_df(relu_result$metrics, "ReLU"),
  build_act_df(tanh_result$metrics, "tanh")
)
p_task2_loss <- ggplot(task2_df %>% filter(metric == "Loss"),
                       aes(x=epoch, y=value, colour=split, linetype=activation)) +
  geom_line(size = 0.8) +
  facet_wrap(~activation) +
  scale_colour_manual(values = c("Train"="#2166ac","Validation"="#d6604d")) +
  labs(title    = "Task 2 – Training & Validation Loss by Activation",
       subtitle = "16 hidden units | lr=0.01 | L2 decay=1e-4",
       x="Epoch", y="Cross-Entropy Loss", colour="Split", linetype="Activation") +
  theme_bw(base_size=13) +
  theme(legend.position="bottom",
        plot.title=element_text(face="bold"),
        strip.background=element_rect(fill="#f0f0f0"))
p_task2_acc <- ggplot(task2_df %>% filter(metric == "Accuracy"),
                      aes(x=epoch, y=value, colour=split, linetype=activation)) +
  geom_line(size=0.8) +
  facet_wrap(~activation) +
  scale_colour_manual(values = c("Train"="#2166ac","Validation"="#d6604d")) +
  labs(title    = "Task 2 – Training & Validation Accuracy by Activation",
       x="Epoch", y="Accuracy", colour="Split", linetype="Activation") +
  theme_bw(base_size=13) +
  theme(legend.position="bottom",
        plot.title=element_text(face="bold"),
        strip.background=element_rect(fill="#f0f0f0"))
ggsave("outputs/task2_loss_comparison.png",     p_task2_loss, width=10, height=5, dpi=150)
ggsave("outputs/task2_accuracy_comparison.png", p_task2_acc,  width=10, height=5, dpi=150)
cat("  [Saved] outputs/task2_loss_comparison.png\n")
cat("  [Saved] outputs/task2_accuracy_comparison.png\n\n")
relu_final_val_acc  <- tail(relu_result$metrics$vl_acc,  1)
tanh_final_val_acc  <- tail(tanh_result$metrics$vl_acc,  1)
relu_final_val_loss <- tail(relu_result$metrics$vl_loss, 1)
tanh_final_val_loss <- tail(tanh_result$metrics$vl_loss, 1)
cat(sprintf("  ReLU final  Val Acc: %.4f  Val Loss: %.4f\n", relu_final_val_acc, relu_final_val_loss))
cat(sprintf("  tanh final  Val Acc: %.4f  Val Loss: %.4f\n\n", tanh_final_val_acc, tanh_final_val_loss))
```

# ── 4. TASK 3 – CAPACITY AND OVERFITTING ──────────────────────────────────────

```{r}
# Lines 421-423
cat("══════════════════════════════════════════════════════════════\n")
cat("  TASK 3: Capacity and Overfitting\n")
cat("══════════════════════════════════════════════════════════════\n\n")
```

# Small network: 8 units

```{r}
# Lines 426-429
cat("  Training Small MLP (8 units)...\n")
small_result <- train_manual_mlp(X_train, y_train, X_val, y_val,
                                 n_hidden=8, act_fn=relu, d_act_fn=d_relu,
                                 lr=0.01, n_epochs=200)
```

# Large network: 128 units (may overfit on 179 training samples)

```{r}
# Lines 432-435
cat("  Training Large MLP (128 units)...\n")
large_result <- train_manual_mlp(X_train, y_train, X_val, y_val,
                                 n_hidden=128, act_fn=relu, d_act_fn=d_relu,
                                 lr=0.01, n_epochs=200)
```

# Large network + L2 regularisation (stronger decay)

```{r}
# Lines 438-441
cat("  Training Large MLP (128 units) + L2 regularisation...\n")
large_l2_result <- train_manual_mlp(X_train, y_train, X_val, y_val,
                                    n_hidden=128, act_fn=relu, d_act_fn=d_relu,
                                    lr=0.01, lambda=1e-2, n_epochs=200)
```

# Dropout MLP ─────────────────────────────────────────────────────────────────
# We implement a dropout-augmented forward pass (training mode drops hidden
# units with probability p_drop; at inference, outputs are scaled by (1-p_drop)).

```{r}
# Lines 447-518
train_dropout_mlp <- function(X_tr, y_tr, X_vl, y_vl,
                              n_hidden   = 128,
                              act_fn     = relu,
                              d_act_fn   = d_relu,
                              lr         = 0.01,
                              lambda     = 1e-4,
                              p_drop     = 0.5,
                              n_epochs   = 200,
                              batch_size = 32,
                              seed       = 42) {
  set.seed(seed)
  n_features <- ncol(X_tr)
  n_classes  <- 2L
  Xtr <- t(as.matrix(X_tr)); Xvl <- t(as.matrix(X_vl))
  ytr <- as.integer(y_tr);   yvl <- as.integer(y_vl)
  W1 <- xavier_init(n_hidden, n_features)
  b1 <- matrix(0, n_hidden, 1)
  W2 <- xavier_init(n_classes, n_hidden)
  b2 <- matrix(0, n_classes, 1)
  n <- ncol(Xtr)
  metrics <- data.frame(epoch=integer(n_epochs), tr_loss=numeric(n_epochs),
                        vl_loss=numeric(n_epochs), tr_acc=numeric(n_epochs),
                        vl_acc=numeric(n_epochs))
  for (ep in seq_len(n_epochs)) {
    perm <- sample(n); Xs <- Xtr[, perm, drop=FALSE]; ys <- ytr[perm]
    starts <- seq(1, n, by=batch_size)
    for (s in starts) {
      ends <- min(s + batch_size - 1, n)
      Xb <- Xs[, s:ends, drop=FALSE]; yb <- ys[s:ends]; nb <- ncol(Xb)
      ## Dropout mask (same mask for all samples in batch for simplicity)
      mask <- rbinom(n_hidden, 1, 1 - p_drop) / (1 - p_drop)
      dW1<-matrix(0,nrow(W1),ncol(W1)); db1<-matrix(0,nrow(b1),1)
      dW2<-matrix(0,nrow(W2),ncol(W2)); db2<-matrix(0,nrow(b2),1)
      for (i in seq_len(nb)) {
        h_pre  <- W1 %*% Xb[, i] + b1
        h      <- act_fn(h_pre) * mask    # apply dropout
        o_pre  <- W2 %*% h + b2
        probs  <- softmax(as.vector(o_pre))
        delta2           <- probs; delta2[yb[i]] <- delta2[yb[i]] - 1
        dh               <- t(W2) %*% delta2
        delta1           <- dh * d_act_fn(h_pre) * mask
        dW2 <- dW2 + delta2 %*% t(h)
        db2 <- db2 + delta2
        dW1 <- dW1 + delta1 %*% t(Xb[, i, drop=FALSE])
        db1 <- db1 + delta1
      }
      W1 <- W1 - lr * (dW1/nb + lambda*W1)
      b1 <- b1 - lr * db1/nb
      W2 <- W2 - lr * (dW2/nb + lambda*W2)
      b2 <- b2 - lr * db2/nb
    }
    ## Evaluation uses full network (no dropout, weights already scaled)
    metrics$epoch[ep]   <- ep
    metrics$tr_loss[ep] <- batch_loss(Xtr, ytr, W1, b1, W2, b2, act_fn, lambda)
    metrics$vl_loss[ep] <- batch_loss(Xvl, yvl, W1, b1, W2, b2, act_fn, lambda)
    metrics$tr_acc[ep]  <- batch_acc(Xtr, ytr, W1, b1, W2, b2, act_fn)
    metrics$vl_acc[ep]  <- batch_acc(Xvl, yvl, W1, b1, W2, b2, act_fn)
  }
  list(W1=W1, b1=b1, W2=W2, b2=b2, metrics=metrics)
}
cat("  Training Large MLP (128 units) + Dropout (p=0.5)...\n")
large_drop_result <- train_dropout_mlp(X_train, y_train, X_val, y_val,
                                       n_hidden=128, n_epochs=200, p_drop=0.5)
```

# ── Plot Task 3 ───────────────────────────────────────────────────────────────

```{r}
# Lines 521-569
build_cap_df <- function(metrics, label) {
  data.frame(
    epoch = rep(metrics$epoch, 4),
    value = c(metrics$tr_loss, metrics$vl_loss, metrics$tr_acc, metrics$vl_acc),
    split = rep(c("Train","Validation","Train","Validation"), each=nrow(metrics)),
    metric = rep(c("Loss","Loss","Accuracy","Accuracy"), each=nrow(metrics)),
    model = label
  )
}
task3_df <- rbind(
  build_cap_df(small_result$metrics,    "Small (8 units)"),
  build_cap_df(large_result$metrics,    "Large (128 units)"),
  build_cap_df(large_l2_result$metrics, "Large + L2"),
  build_cap_df(large_drop_result$metrics, "Large + Dropout")
)
task3_df$model <- factor(task3_df$model,
                         levels = c("Small (8 units)","Large (128 units)",
                                    "Large + L2","Large + Dropout"))
p_task3_loss <- ggplot(task3_df %>% filter(metric=="Loss"),
                       aes(x=epoch, y=value, colour=split)) +
  geom_line(size=0.7) +
  facet_wrap(~model, nrow=2) +
  scale_colour_manual(values=c("Train"="#2166ac","Validation"="#d6604d")) +
  labs(title="Task 3 – Capacity & Regularisation: Loss Curves",
       x="Epoch", y="Cross-Entropy Loss", colour="Split") +
  theme_bw(base_size=12) +
  theme(legend.position="bottom",
        plot.title=element_text(face="bold"),
        strip.background=element_rect(fill="#f0f0f0"))
p_task3_acc <- ggplot(task3_df %>% filter(metric=="Accuracy"),
                      aes(x=epoch, y=value, colour=split)) +
  geom_line(size=0.7) +
  facet_wrap(~model, nrow=2) +
  scale_colour_manual(values=c("Train"="#2166ac","Validation"="#d6604d")) +
  labs(title="Task 3 – Capacity & Regularisation: Accuracy Curves",
       x="Epoch", y="Accuracy", colour="Split") +
  theme_bw(base_size=12) +
  theme(legend.position="bottom",
        plot.title=element_text(face="bold"),
        strip.background=element_rect(fill="#f0f0f0"))
ggsave("outputs/task3_loss_capacity.png",     p_task3_loss, width=12, height=8, dpi=150)
ggsave("outputs/task3_accuracy_capacity.png", p_task3_acc,  width=12, height=8, dpi=150)
cat("  [Saved] outputs/task3_loss_capacity.png\n")
cat("  [Saved] outputs/task3_accuracy_capacity.png\n\n")
```

# Summary table

```{r}
# Lines 572-587
t3_metrics <- lapply(
  list(Small   = small_result,
       Large   = large_result,
       LargeL2 = large_l2_result,
       LargeDO = large_drop_result),
  function(r) {
    m <- r$metrics
    c(final_tr_acc  = round(tail(m$tr_acc,  1), 4),
      final_vl_acc  = round(tail(m$vl_acc,  1), 4),
      final_tr_loss = round(tail(m$tr_loss, 1), 4),
      final_vl_loss = round(tail(m$vl_loss, 1), 4))
  }
)
cat("── Task 3 Final Metrics ──\n")
print(do.call(rbind, t3_metrics))
cat("\n")
```

# ── 5. TASK 4 – RESPONSIBLE EVALUATION (single test-set evaluation) ──────────

```{r}
# Lines 591-593
cat("══════════════════════════════════════════════════════════════\n")
cat("  TASK 4: Responsible Evaluation — Final Test-Set Report\n")
cat("══════════════════════════════════════════════════════════════\n\n")
```

# We finalise model selection based SOLELY on validation metrics, then call
# the test set exactly once per model.

# Helper for manual MLP prediction

```{r}
# Lines 599-653
predict_manual <- function(result, X, act_fn = relu) {
  W1 <- result$W1; b1 <- result$b1; W2 <- result$W2; b2 <- result$b2
  Xm <- t(as.matrix(X))
  probs_died <- sapply(seq_len(ncol(Xm)), function(i) {
    fp <- forward(Xm[, i], W1, b1, W2, b2, act_fn)
    fp$probs[2]   # P(Died)
  })
  probs_died
}
eval_final <- function(prob, y_true, model_name) {
  pred    <- factor(ifelse(prob >= 0.5, "Died", "Survived"),
                    levels = c("Survived", "Died"))
  acc     <- mean(pred == y_true)
  cm      <- confusionMatrix(pred, y_true, positive = "Died")
  roc_obj <- roc(as.numeric(y_true == "Died"), prob, quiet = TRUE)
  auc_val <- as.numeric(auc(roc_obj))
  sens    <- cm$byClass["Sensitivity"]
  spec    <- cm$byClass["Specificity"]
  cat(sprintf("  %-35s  Acc: %.4f  AUC: %.4f  Sens: %.4f  Spec: %.4f\n",
              model_name, acc, auc_val, sens, spec))
  invisible(list(acc=acc, auc=auc_val, sens=sens, spec=spec))
}
cat("── Logistic Regression (test) ──\n")
lr_prob_test <- predict(lr_model, newdata=X_test, type="response")
eval_final(lr_prob_test, y_test, "Logistic Regression")
cat("\n── MLP: Task 1 (nnet, sigmoid, 16u) (test) ──\n")
mlp_prob_test <- as.vector(predict(mlp_model_task1, newdata=X_test, type="raw"))
eval_final(mlp_prob_test, y_test, "MLP Task1 (sigmoid, 16u)")
cat("\n── Manual MLP: ReLU 16u ──\n")
relu_prob_test <- predict_manual(relu_result, X_test, relu)
eval_final(relu_prob_test, y_test, "ReLU MLP (16u)")
cat("\n── Manual MLP: tanh 16u ──\n")
tanh_prob_test <- predict_manual(tanh_result, X_test, tanh_fn)
eval_final(tanh_prob_test, y_test, "tanh MLP (16u)")
cat("\n── Manual MLP: Small (8u, ReLU) ──\n")
small_prob_test <- predict_manual(small_result, X_test, relu)
eval_final(small_prob_test, y_test, "Small MLP (8u, ReLU)")
cat("\n── Manual MLP: Large (128u, ReLU) ──\n")
large_prob_test <- predict_manual(large_result, X_test, relu)
eval_final(large_prob_test, y_test, "Large MLP (128u, ReLU)")
cat("\n── Manual MLP: Large + L2 (128u) ──\n")
l2_prob_test <- predict_manual(large_l2_result, X_test, relu)
eval_final(l2_prob_test, y_test, "Large MLP + L2 (128u)")
cat("\n── Manual MLP: Large + Dropout (128u) ──\n")
do_prob_test <- predict_manual(large_drop_result, X_test, relu)
eval_final(do_prob_test, y_test, "Large MLP + Dropout (128u)")
```

# ── 6. COMBINED SUMMARY BAR CHART ────────────────────────────────────────────

# Validation AUC summary (used for selection decisions)

```{r}
# Lines 658-695
val_auc_summary <- data.frame(
  Model = c("Logistic Reg.", "MLP (sigmoid,16u)",
            "ReLU 16u", "tanh 16u",
            "Small (8u)", "Large (128u)",
            "Large+L2", "Large+Dropout"),
  Val_AUC = c(lr_val$auc,
              mlp_val$auc,
              auc(roc(as.numeric(y_val=="Died"),
                      predict_manual(relu_result,  X_val, relu),    quiet=TRUE)),
              auc(roc(as.numeric(y_val=="Died"),
                      predict_manual(tanh_result,  X_val, tanh_fn), quiet=TRUE)),
              auc(roc(as.numeric(y_val=="Died"),
                      predict_manual(small_result, X_val, relu),    quiet=TRUE)),
              auc(roc(as.numeric(y_val=="Died"),
                      predict_manual(large_result, X_val, relu),    quiet=TRUE)),
              auc(roc(as.numeric(y_val=="Died"),
                      predict_manual(large_l2_result,   X_val, relu), quiet=TRUE)),
              auc(roc(as.numeric(y_val=="Died"),
                      predict_manual(large_drop_result, X_val, relu), quiet=TRUE)))
)
val_auc_summary$Model <- factor(val_auc_summary$Model,
                                 levels = rev(val_auc_summary$Model))
p_summary <- ggplot(val_auc_summary, aes(x=Model, y=Val_AUC, fill=Val_AUC)) +
  geom_col(width=0.65, colour="white") +
  geom_text(aes(label=sprintf("%.3f", Val_AUC)), hjust=-0.1, size=3.8) +
  coord_flip(ylim=c(0.55, 1.0)) +
  scale_fill_gradient(low="#abd9e9", high="#2166ac") +
  labs(title="Summary – Validation AUC Across All Models",
       subtitle="Higher is better | Used for model selection only",
       x=NULL, y="ROC-AUC (Validation)") +
  theme_bw(base_size=13) +
  theme(legend.position="none",
        plot.title=element_text(face="bold"))
ggsave("outputs/summary_val_auc.png", p_summary, width=9, height=5, dpi=150)
cat("\n  [Saved] outputs/summary_val_auc.png\n")
```
# Conclusion
Feedforward neural networks on the Heart Failure Clinical Records dataset, the more space
the architecture has, the less the generalisation on small clinical datasets. Both the logistic
regression and the one hidden-layer MLP have similar test AUC of 0.81, which indicates that
the use of nonlinearity is only marginally beneficial under the constraints of the data. Tanh
has a lower convergence rate than ReLU, and the unregularised large network has a high
degree of overfitting. Principled regularisation and disciplined train-validation-test separation
are the two most important features in ensuring reliable medical machine learning evaluation,
which can be achieved through dropout and L2 regularization to restore validation stability.
